---
title: "Milestone 0: Rwanda district-level VCI smoke test"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Milestone 0: Rwanda district-level VCI smoke test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Only build the figures if the suggested plotting packages are available.
buildable <- requireNamespace("ggplot2", quietly = TRUE) &&
  requireNamespace("sf", quietly = TRUE) &&
  requireNamespace("patchwork", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = buildable,
  fig.align = "center",
  fig.width = 7.2,
  fig.height = 4.3,
  dpi = 300,
  out.width = "100%"
)
```

> **⚠️ Illustrative only — not a real analysis.**
> This vignette is the Milestone 0 *smoke test*: an end-to-end run built to shake
> out architecture and integration problems *before* the science is signed off.
> It uses ballpark parameters and a deliberately crude, clearly-labelled stub for
> the intervention effects. **Its numbers must not be read as findings about
> malaria, vectors, or bed nets in Rwanda.**

## What this computes

Vector control impact (VCI) — the proportional reduction in vectorial capacity —
across the districts (GADM admin-2) of Rwanda, comparing three scenarios:

- **No intervention** — the reference;
- **Business-as-usual (BAU)** — current net use, assumed all **pyrethroid** LLIN;
- **Alternative** — the same net use, but all nets replaced with **dual-AI** LLIN.

The inputs are the packaged [`rwanda_inputs`] dataset: district-level abundance
for the three primary vectors (*An. gambiae*, *An. arabiensis*, *An. funestus*),
pyrethroid susceptibility, and 2024 net use, with ballpark bionomics.

```{r setup}
library(vci)

inputs <- rwanda_inputs
inputs
```

```{r plot-helpers}
library(ggplot2)
library(sf) # for the geometry merge and geom_sf()
library(patchwork) # for side-by-side panels with independent colour scales

districts <- inputs$geometry
near_white <- "#f7f7f7"

# A clean choropleth of one value column, on a near-white -> `high` gradient.
district_map <- function(data, column, low, high, legend,
                         percent = FALSE, limits = NULL, breaks = waiver()) {
  labels <- if (percent) function(x) paste0(x, "%") else waiver()
  ggplot(merge(districts, data, by = "location_id")) +
    geom_sf(aes(fill = .data[[column]]), colour = "grey70", linewidth = 0.12) +
    scale_fill_gradient(
      low = low, high = high, name = legend,
      labels = labels, limits = limits, breaks = breaks
    ) +
    theme_void(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5, margin = margin(b = 6)),
      strip.text = element_text(face = "bold", margin = margin(t = 4, b = 4)),
      legend.position = "right"
    )
}
```

## Input data

### Vector abundance by species

Adults per human (`m`) for each of the three vectors, on a shared scale so their
magnitudes are comparable. *An. gambiae* dominates the west, *An. arabiensis* the
east and south; *An. funestus* is scarcer throughout.

```{r fig-abundance, fig.width = 8.2, fig.height = 4}
vectors <- inputs$vectors
abundance_max <- max(vectors$abundance)

species_map <- function(sp, high) {
  d <- vectors[vectors$species == sp, c("location_id", "abundance")]
  district_map(d, "abundance",
    low = near_white, high = high,
    legend = "mosquitoes\nper human", limits = c(0, abundance_max)
  ) +
    labs(title = sp) +
    theme(legend.position = "bottom")
}

(species_map("gambiae", "#2ca25f") |
  species_map("arabiensis", "#cb181d") |
  species_map("funestus", "#2171b5")) +
  plot_annotation(
    title = "Vector abundance by species",
    theme = theme(plot.title = element_text(face = "bold", hjust = 0.5))
  )
```

### Intervention context

Net use and insecticide susceptibility, both on a 0–100% scale and coloured so
that **near-white means less protection**: low net use, or low susceptibility
(i.e. high resistance).

```{r fig-context, fig.width = 8, fig.height = 4}
sites_df <- inputs$sites
sites_df$net_use_pct <- 100 * sites_df$net_use
sites_df$susceptibility_pct <- 100 * sites_df$susceptibility

p_llin <- district_map(sites_df, "net_use_pct",
  low = near_white, high = "#08519c",
  legend = "LLIN use", percent = TRUE,
  limits = c(0, 100), breaks = c(0, 25, 50, 75, 100)
) +
  labs(title = "LLIN use (2024)")

p_susc <- district_map(sites_df, "susceptibility_pct",
  low = near_white, high = "#6a51a3",
  legend = "Susceptibility", percent = TRUE,
  limits = c(0, 100), breaks = c(0, 25, 50, 75, 100)
) +
  labs(title = "Pyrethroid susceptibility")

p_llin | p_susc
```

## Running the three scenarios

`compute_capacity()` runs the whole pipeline — intervention stub, host choice,
survival, the capacity equation, and the species sum — for each scenario, giving
total vectorial capacity per district.

```{r compute}
capacity <- compute_capacity(inputs)
head(capacity)
```

`vci_by_scenario()` then compares each scenario against a named reference within
each district.

```{r vci}
vci_vs_none <- vci_by_scenario(capacity, reference = "none")
vci_vs_bau <- vci_by_scenario(capacity, reference = "bau")

scenario_labels <- c(
  none = "No intervention",
  bau = "BAU: pyrethroid LLIN",
  alt = "Alternative: dual-AI LLIN"
)
label_scenarios <- function(x) factor(scenario_labels[x], levels = scenario_labels)
```

## Results

### Vectorial capacity under each scenario

Both interventions pull capacity down relative to no intervention, and the
dual-AI nets pull it down furthest.

```{r fig-capacity, fig.width = 7.6, fig.height = 3.8}
capacity$panel <- label_scenarios(capacity$scenario)

district_map(capacity, "capacity",
  low = near_white, high = "#b30000",
  legend = "Vectorial\ncapacity", limits = c(0, max(capacity$capacity))
) +
  facet_wrap(~panel) +
  labs(title = "Vectorial capacity by district") +
  theme(plot.title = element_text(
    face = "bold", hjust = 0.5, margin = margin(b = 16)
  ))
```

### VCI of each net against no intervention

Vector control impact as a percentage, on a fixed 0–100% scale. The dual-AI panel
is much darker: with Rwanda's high pyrethroid resistance, pyrethroid killing is
weak, while the dual-AI's resistance-independent killing keeps working.

```{r fig-vci-vs-none, fig.width = 7.4, fig.height = 3.8}
vci_pct <- vci_vs_none[vci_vs_none$scenario != "none", ]
vci_pct$vci_percent <- 100 * vci_pct$vci
vci_pct$panel <- label_scenarios(vci_pct$scenario)

district_map(vci_pct, "vci_percent",
  low = near_white, high = "#00897b",
  legend = "VCI (%)", percent = TRUE,
  limits = c(0, 100), breaks = c(0, 25, 50, 75, 100)
) +
  facet_wrap(~panel) +
  labs(title = "Vector control impact vs no intervention")
```

### Added impact of dual-AI over business-as-usual

Comparing the alternative directly against BAU isolates the extra impact of
switching pyrethroid nets for dual-AI nets, at the same coverage.

```{r fig-vci-alt-vs-bau, fig.width = 5.8, fig.height = 4.4}
vci_alt <- vci_vs_bau[vci_vs_bau$scenario == "alt", ]
vci_alt$vci_percent <- 100 * vci_alt$vci

district_map(vci_alt, "vci_percent",
  low = near_white, high = "#00897b",
  legend = "VCI (%)", percent = TRUE,
  limits = c(0, 100), breaks = c(0, 25, 50, 75, 100)
) +
  labs(title = "Extra impact of dual-AI over pyrethroid (alt vs BAU)")
```

## Notes on the assumptions

- The **intervention effects are a stub** (`intervention_effect_stub()`):
  plausible round numbers, not a fitted mapping. The one qualitatively real
  feature is that pyrethroid killing scales with susceptibility (it collapses as
  resistance rises), whereas dual-AI adds resistance-independent killing.
- **Host choice.** Host-opportunity weights are derived from the
  malariasimulation `Q0` (human blood index) and `phi_indoors` under an explicit
  *equal-host-availability* assumption — `Q0` is treated as preference. This is a
  simplification adopted deliberately for the smoke test, not a modelling
  commitment.
- **Abundance cancels.** Because the interventions here change survival and
  feeding but not abundance, absolute abundance cancels out of VCI. Only the
  capacity maps depend on it; the VCI maps do not.
```
